---
title: "Coding Physical Intimate Partner Violence in Suicide Narratives with a Local LLM: A step-by-step, transparent walkthrough of the coding pipeline"
---

```{r setup, include=FALSE}
# This document is a TRANSPARENCY WALKTHROUGH, not an executable analysis.
# Code chunks are shown for documentation but are not run when the file is knit
# (eval = FALSE), because the underlying narratives are restricted data and the
# model is served locally. The goal is to let a reader follow exactly what the
# procedure does, step by step.
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE)
```

## What this document is

This is a complete, readable record of how we use a **locally hosted large
language model (LLM)** to code suicide case narratives for evidence of
**physical intimate partner violence (IPV)** in the period before the death.

It is written so that another researcher can follow every step and understand
exactly what the model is asked, how its answers are captured, and how the
results are assembled — *without* needing to run anything and *without* access
to the underlying data.

A few framing points before the code:

- **Why a local model.** The narratives are sensitive, restricted-access
  records. We run an open-weight model (`llama3.2:3b`) locally through
  [Ollama](https://ollama.com), so the text never leaves the analysis machine
  and is never sent to a third-party API.
- **Two narratives per case.** Each case has two separate source reports: a
  **law enforcement (LE)** narrative and a **coroner/medical examiner (ME)**
  narrative. They describe the decedent and the circumstances before the death
  from two different vantage points.
- **Each report is coded on its own.** We deliberately send one narrative to
  the model at a time. The model never sees both reports together, so a code
  for the LE narrative cannot be influenced by the ME narrative or vice versa.
- **Determinism.** The model is queried with `temperature = 0`, which makes its
  output as reproducible as the model allows for a given prompt.

---

## Background: how this fits into the larger study

The script below is one piece of a larger study. This section explains the
reasoning behind it — the parts a reader needs to understand the method but that
the code itself does not show — and mirrors what is reported in the manuscript.

### The data and the task

The narratives come from the National Violent Death Reporting System (NVDRS),
covering 2003–2019. Every suicide case in this data includes **two** narratives
written by different sources: a **law enforcement (LE)** report and a
**coroner/medical examiner (ME)** report. Each describes the person's background
and the circumstances leading up to the death.

The goal here is narrow and specific: decide whether **physical** intimate
partner violence appears in the circumstances before the suicide, and record it
as a simple yes/no (1 or 0). We focused on *physical* IPV — rather than emotional
or sexual IPV — because physical violence usually shows up as concrete,
observable actions (someone was hit, choked, and so on). That makes it easier to
define clearly and to code consistently than more subjective forms of abuse.

### A human-coded "answer key"

Before involving any model, the research team hand-coded the cases to create a
gold-standard reference set — the yardstick the model is later measured against.
Working from an existing coding framework, the team built a codebook, refined it
over three rounds of pilot testing on a random sample of 120 cases, and resolved
disagreements through discussion until they reached full agreement. Two coders
then independently coded all cases, and two additional team members reviewed the
initial results and any unclear cases. In the end, **86 cases (11.1%)** were
identified as involving physical IPV before the suicide. Every model's output is
ultimately compared back to this human-coded set.

### Why four models, shown as one

This walkthrough shows the pipeline for a single model, but the study actually
ran **four** small, open-weight models from four different developers:

- **Llama 3.2** (3 billion parameters)
- **Gemma 3** (4B)
- **Qwen 2.5** (3B)
- **Phi-3 mini** (3.8B)

We deliberately chose small models (all under 4 billion parameters) for
practical reasons. Despite their size, they still handle language well enough to
interpret context and follow instructions; they are small enough to run on an
ordinary personal laptop, with no high-end GPU required; and using models from
different organizations lets us compare their performance and later combine them.
This matters for reproducibility: another researcher without specialized
computing equipment can repeat the approach.

The code is **identical for every model** — only the model's name changes — so it
is shown once here rather than repeated four times.

### Running everything locally

All four models were downloaded with [Ollama](https://ollama.com) and then run
completely offline. An internet connection was needed only to download the
models the first time; after that, no case data ever left the machine. All
processing happened on a secure, password-protected standalone laptop, so the
sensitive narratives stayed entirely within the local environment. (This is why
the code talks to a model at `localhost`, not a web address.)

### How the prompt was developed

The instruction block in Step 4 is the **final** version of the prompt, and it
did not start out that way. The first prompt was more elaborate: it used a
"few-shot" style (a handful of worked coding examples), included definitions and
the acronyms that appear in the narratives, and asked the model to produce a
single combined code for each case.

We refined the prompt over **three rounds of pilot testing on about 10% of the
data (n = 75)**:

1. **First, on 10 cases.** The initial prompt produced a lot of
   misclassification, so it needed substantial revision.
2. **Then, on 15 more cases**, using the revised prompt, with small wording
   tweaks afterward — for example, adding the instruction *"Code 1 ONLY when
   there is clear and explicit evidence of a physical violent act."*
3. **Finally, on 50 more cases**, to lock in the standardized prompt used for the
   full dataset.

Two lessons from that testing shaped the final prompt. First, **shorter and more
direct worked better**: the worked examples and acronym list were removed to
reduce complexity, and wording drawn from the NVDRS narratives themselves was
used instead. Second, we **simplified the unit of coding**: instead of asking the
model to merge both narratives into one case-level decision, the final prompt
asks it to code each report (LE and ME) **separately**. That second change is
exactly the design this walkthrough follows.

---

## Step 1 — Packages

The pipeline relies on a small set of packages: `httr` (HTTP requests to the
local model), `jsonlite` (JSON handling), `readxl` / `writexl` (Excel I/O),
`stringr` (text parsing), and `dplyr` (data manipulation).

```{r install, eval=FALSE}
# Run once, if these are not already installed.
install.packages(c("httr", "jsonlite", "readxl", "stringr", "dplyr", "writexl"))
```

```{r libraries}
library(httr)
library(jsonlite)
library(readxl)
library(stringr)
library(dplyr)
library(writexl)
```

---

## Step 2 — Working directory and project folder

We point R at the project folder. Everything else (the data file, the output
folder) is referenced *relative* to this location, so the script is portable
across machines as long as the folder layout is the same.

```{r setwd}
setwd("~/Documents/R/LLM sample")
getwd()  # confirm we are where we expect to be

# Create an output folder if it does not already exist.
if (!dir.exists("output")) dir.create("output")
```

---

## Step 3 — Read the data

The input is an Excel file already structured in **long format** — that is, one
row per *narrative* rather than one row per case. Because each case has two
narratives (LE and ME), a single case appears across two rows. The file has the
following columns:

| Column        | Meaning                                  |
|---------------|------------------------------------------|
| `case_id`     | Case identifier                          |
| `report_type` | Which narrative the row is (`LE` or `ME`)|
| `report`      | The narrative text to be coded           |

Storing the data this way means **no reshaping is needed**: every row already
holds a single narrative, tagged with the case it belongs to and which source it
came from, which is exactly what lets the model code one report at a time.
Reading the file is therefore a single step.

```{r read-data}
data_long <- read_excel("data/LLM sample.xlsx")

# Make sure the key columns are plain text (character), so the prompt and the
# loops downstream behave consistently. Adjust the column names on the left if
# yours differ.
data_long <- data_long %>%
  mutate(
    case_id     = as.character(case_id),
    report_type = as.character(report_type),
    report      = as.character(report)
  )

head(data_long)
```

If your column names differ from `case_id`, `report_type`, and `report`, either
rename them when you read the file or update the names used in the rest of the
script so they match.

---

## Step 4 — The coding prompt

This is the heart of the method and the part most worth scrutinizing. The
function below wraps a single narrative in a fixed instruction block. **Every
narrative receives exactly this prompt**, with only the narrative text changing.

The instructions encode a deliberately conservative coding standard:

- **Default to 0.** The model is told to assume *no* physical IPV unless the
  text gives explicit evidence otherwise.
- **Explicit acts only.** A code of 1 requires a clearly described physical act
  (e.g., hit, choke, stab) between intimate partners — not inference, conflict,
  or vague mentions of "abuse."
- **Timing matters.** Violence is coded 1 only when it is recent or connected to
  the circumstances of the death; distant, unconnected history is coded 0.
- **Who is an intimate partner** is defined narrowly (partners and ex-partners,
  not other relatives).
- **Role-neutral.** The decedent may be victim, perpetrator, or both; the code
  reflects whether partner violence occurred, not who initiated it.
- **Output is a bare number.** The model is asked to return only `0` or `1`,
  with no labels or explanation, which keeps parsing simple and unambiguous.

```{r build-prompt}
build_prompt <- function(report_text) {
  paste0(
"You are coding suicide case narratives for evidence of physical intimate partner violence (IPV) prior to the suicide.

Task:
Determine whether physical intimate partner violence is present (1) or not present (0).

Important coding standard:
- Default to 0.
- Code 1 ONLY when there is clear and explicit evidence of a physical violent act.
- Do NOT infer, assume, or guess.
- If the evidence is vague, indirect, or uncertain, code 0.

Definitions:
Intimate partner:
Spouse, fiance, boyfriend/girlfriend, dating partner, ex-partner, or significant other.
Do NOT include family members or other relatives.

Physical violence:
Code 1 ONLY if the narrative explicitly describes a physical act that causes or attempts to cause bodily harm between intimate partners.

Valid examples:
Physically abusive, hit, slap, punch, push, choke, kick, beat, stab, shoot, burn, physically assault.

Timing rule:
- Code 1 only if the physical violence appears to be recent or related to the circumstances before the suicide.
- If the violence is described only as a distant or past event with no clear connection to suicide, code 0.

NOT physical violence:
- No known physical altercation
- Harm themselves
- arguments
- relationship conflict
- divorce or separation
- restraining orders alone
- jealousy
- verbal fights
- emotional distress
- vague mention of abuse without specific physical harm
- past violence not connected to suicide

If you are unsure, code 0.

Role clarification:
The decedent may be a victim, a perpetrator, or both.
Code based on whether physical violence occurred between partners, regardless of who initiated it.

Return ONLY a single number:
0 = no physical intimate partner violence
1 = physical intimate partner violence present
Do not include any text, labels, or explanation.

Narrative:
", report_text)
}
```

---

## Step 5 — Querying the local model

This function sends a prompt to the Ollama server running on the same machine
(`localhost:11434`) and returns the model's text response. Setting
`temperature = 0` requests the most deterministic output, and `stream = FALSE`
returns the full answer in a single response object.

```{r query-ollama}
query_ollama <- function(prompt_text, model_name = "llama3.2:3b") {
  res <- POST(
    url  = "http://localhost:11434/api/generate",
    body = list(
      model   = model_name,
      prompt  = prompt_text,
      stream  = FALSE,
      options = list(temperature = 0)  # deterministic
    ),
    encode = "json"
  )
  stop_for_status(res)
  content(res, as = "parsed", type = "application/json")$response
}
```

The pipeline is shown with the selected model, but the **same `query_ollama()`
function was used for all four models** during the study — only the `model_name`
argument changed (for example, `"gemma3"`, `"qwen2.5:3b"`, or `"phi3:mini"`).
That is why a single version of the code is enough to document the whole
procedure.

---

## Step 6 — Parsing the model's answer

The prompt asks for a bare `0` or `1`, so the parser simply extracts the first
`0`/`1` from the (trimmed) response and returns it as an integer. If nothing
usable is found, it returns `NA` rather than guessing.

Crucially, we keep the model's **raw output** alongside the parsed code (see the
loops below). That makes every decision auditable: if a parsed value ever looks
wrong, the original text is right there to check against.

```{r parse-output}
parse_output <- function(text) {
  m <- str_extract(str_trim(text), "[01]")
  data.frame(Physical = ifelse(is.na(m), NA_integer_, as.integer(m)))
}
```

---

## Step 7 — Test a single case first

Before running anything at scale, we inspect one prompt in full and look at the
model's answer. This confirms the prompt reads the way we intend and that the
model returns a clean, parseable response.

```{r test-one}
test_prompt <- build_prompt(data_long$report[1])
cat(test_prompt)

test_result <- query_ollama(test_prompt)
cat(test_result)
parse_output(test_result)
```

---

## Step 8 — Run a small batch as a sanity check

Next we code just the first five rows. This is a guard against silently
applying a flawed prompt or a misconfigured model to the entire dataset: it is
much cheaper to catch a problem on five rows than on the full file.

Each iteration builds the prompt, queries the model (wrapped in `tryCatch` so a
single failure records an error string instead of halting the whole run), parses
the answer, and stores the case ID, report type, raw output, and parsed code
together.

```{r small-batch}
small_data   <- data_long[1:5, ]
results_list <- vector("list", nrow(small_data))

for (i in seq_len(nrow(small_data))) {
  cat("Running row", i, "of", nrow(small_data), "\n")

  prompt_i <- build_prompt(small_data$report[i])

  raw_result <- tryCatch(
    query_ollama(prompt_i),
    error = function(e) paste("ERROR:", e$message)
  )

  parsed_result <- parse_output(raw_result)

  results_list[[i]] <- cbind(
    data.frame(
      case_id     = small_data$case_id[i],
      report_type = small_data$report_type[i],
      raw_output  = raw_result,
      stringsAsFactors = FALSE
    ),
    parsed_result
  )
}

small_results <- bind_rows(results_list)
small_results
```

Once these five rows look right — sensible codes, raw outputs that match the
codes — we proceed to the full dataset.

---

## Step 9 — Run the full dataset

The full run is identical in structure to the small batch, just over every row
of `data_long`. The result is one row per narrative, each carrying its case ID,
report type (LE or ME), the model's raw text, and the parsed `Physical` code.

```{r full-run}
results_list <- vector("list", nrow(data_long))

for (i in seq_len(nrow(data_long))) {
  cat("Running row", i, "of", nrow(data_long), "\n")

  prompt_i <- build_prompt(data_long$report[i])

  raw_result <- tryCatch(
    query_ollama(prompt_i),
    error = function(e) paste("ERROR:", e$message)
  )

  parsed <- parse_output(raw_result)

  results_list[[i]] <- cbind(
    data.frame(
      case_id     = data_long$case_id[i],
      report_type = data_long$report_type[i],
      raw_output  = raw_result,
      stringsAsFactors = FALSE
    ),
    parsed
  )
}

results_long <- bind_rows(results_list)
```

---

## Step 10 — Save the results

Finally we write the coded results to the `output/` folder. The saved file keeps
the raw model output next to each code, so the results remain fully auditable
after the fact.

```{r save}
write_xlsx(results_long, "output/results_llama3.2_3b.xlsx")
```

If you want one code per *case* rather than per report, you can collapse the two
rows afterward (for example, flagging a case as 1 if either its LE or ME code is
1) — but note that this combining rule is then applied transparently in R,
rather than being left to the model.

---

## Beyond this script: case-level coding, validation, and ensembles

The code above stops where one model finishes coding every narrative. In the
full study, three further steps follow — described here in plain language so the
walkthrough matches the manuscript, even though they are not the focus of this
script.

### From two report codes to one case code

Because each report (LE and ME) is coded separately, every case ends up with two
codes. The team combined them into a single **case-level** code using a simple
rule: a case counts as physical IPV (1) if **either** its LE report **or** its ME
report was coded 1; otherwise it is 0. This combining step is done by the
research team in R — not by the model — so the rule is explicit and consistent.

### Checking the codes against the human "answer key"

Each model's case-level codes were then compared with the human-coded reference
set to see how well the model did. Five measures were used:

- **Accuracy** — the share of all cases the model classified correctly.
- **Recall (sensitivity)** — of the cases that truly involved physical IPV, the
  share the model successfully caught.
- **Precision** — of the cases the model flagged as physical IPV, the share that
  truly were.
- **F1 score** — a single balanced number that combines precision and recall.
- **Cohen's kappa** — how much the model and the human coders agreed, adjusted
  for the agreement you would expect by chance alone.

### Combining models (ensembles)

Finally, because the four models had somewhat different strengths, their outputs
were combined to see whether the group did better than any single model. Two
strategies were tested: a **majority vote** (use whichever code most of the four
models chose) and a **performance-based rule** (when the models disagreed, give
more weight to the models that had scored better on the F1 score and Cohen's
kappa).

---

## Transparency and reproducibility notes

- **Exact prompt.** The instruction block in Step 4 is applied verbatim to every
  narrative; nothing is added or removed per case.
- **Model and settings.** Results were produced with `llama3.2:3b` served
  locally via Ollama at `temperature = 0`. Different models, model versions, or
  temperatures can change the output, so these are reported explicitly.
- **Raw outputs retained.** Every parsed code is stored next to the model's
  original response, so any code can be checked against what the model actually
  said.
- **Known limitations.** An LLM may occasionally deviate from the requested
  format or apply the coding standard imperfectly. A human audit of a sample of
  cases — comparing `raw_output` to the parsed `Physical` code — is recommended
  before treating the codes as final.
- **Data access.** The narratives are restricted and are not distributed with
  this document. Only the procedure is shared here; reproducing the numbers
  requires separately authorized access to the source data.
